The promise of community-driven preprints in ecology and evolution

### 2024-05-01 ms/ms.qmd
Add a few refs and fix up intro

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While preprint use in biology has had a slower uptake [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints are expected to promote open science and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings; 2) helping establish precedence of research findings; 3) removing financial barriers to publication; and 4) encouraging constructive and critical feedback from the research community [@Proulx2013; @bourne2017ten; @vale2015accelerating].

Preprint servers can empower researchers to make their findings more accessible, open, and transparent but only if they are used well and initiatives are widely known. Significant barriers to preprint use still exist globally – from lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by a research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution which can help inform future initiatives, improve others, and allow us to work harder in further breaking down barriers to their use in our field.

*EcoEvoRxiv* is one of the few community driven preprint servers that has paved the way for new initiatives, including accepting multilingual preprints, recognising and accepting registered reports and non-traditional research reports, and promoting peer review and community discussion to improve the quality of preprints and speed up their peer-reviewed publication (e.g., Peer Community In -- PCI).

Here, we explore the first preprints uploaded to *EcoEvoRxiv* to understand preprint practices and use in ecology and evolution. We focus our attention on understanding better: 1) what countries authors that preprint are located; 2) how career stage and gender impact preprint use; 3) the extent to which authors make use of preprint servers for registered reports and community-driven peer review; 4) whether data and code are more likely to be shared in preprints; and 5) how long it takes for a preprint to become published, and how many preprints remain unpublished. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other servers which we think will help further educate researchers about the benefits of using community driven preprint servers to post their research findings.

>>>>

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While preprint use in biology has had a slower uptake [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints are expected to promote open science, increase the visibility of research (e.g., [fu2019releasing]), and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings; 2) helping establish precedence of research findings; 3) removing financial barriers to publication; and 4) encouraging constructive and critical feedback from the research community [@Proulx2013; @bourne2017ten; @vale2015accelerating].

Preprint servers can empower researchers to make their findings more accessible, open, and transparent but only if they are used as forums for spreading and discussing findings within a research community. Significant barriers to preprint use still exist globally – from a lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by the research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution that can help inform future initiatives and allow us to work harder in further breaking down barriers to preprint use in our field.

*EcoEvoRxiv* is one of the few community-driven preprint servers that has paved the way for new initiatives, including by accepting multilingual preprints, recognising registered reports and non-traditional research reports, and promoting peer review and community discussion to improve the quality of preprints and speed up their peer-reviewed publication (e.g., Peer Community In -- PCI [@guillemaud2019peer]). Peer Community In allows for fast constructive peer review around a preprint. Peer reviews are transparent and published online and the community makes a recommendation on where the reasearch findings should be published [@guillemaud2019peer].

Here, we explore the first preprints uploaded to *EcoEvoRxiv* to understand preprint practices and uses in ecology and evolution. We focus our attention on better understanding: 1) what countries authors who use *EcoEvoRxiv* are located; 2) how career stage and gender impact preprint server use; 3) the extent to which authors make use of preprint servers for registered reports and community-driven peer review; 4) the extent to which data and code are shared in preprints; and 5) how many preprints remain unpublished, and for those that are published, how long it took them to become published. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other preprint servers to help further educate researchers about the benefits of using community-driven preprint servers to post their research findings.
